\documentclass[a4paper,12pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{listings}
\title{Manglang}
\author{Magnus Burenius}
\begin{document}

\maketitle

\section{Overiew}

Mang Lang is a functional language without any side efects. The lifetime of a Mang Lang program has four stages:
\begin{enumerate}
\item Program written in Mang Lang as text.
\item Program parsed into a tree graph.
\item Program evaluated into another tree graph.
\item Program result presented in Mang Lang as text.
\end{enumerate}

\section{Minimal Language for Data}

We start by defining a minimal language for storing data, similar to Json. It consists of these primitive values:
\begin{itemize}
\item Numbers like: \lstinline|1.0|
\item Characters like: \lstinline|'a'|
\end{itemize}
Values can be packaged in these collections:
\begin{itemize}
\item Dictionaries like: \lstinline|{x = 1  y = 2}|
\item Lists like: \lstinline|(1 2 3)|
\item Strings like: \lstinline|''Magnus''|
\end{itemize}
Dictionaries and strings can be nested.

\section{Extended Language for Data}

We add the posibility to refer back to symbols:
\begin{itemize}
\item Reference to symbol in current scope like: \lstinline|{x = 1 y = x}|
\item Reference to symbol in environment scope like:  \lstinline|{x = 1 y = {z = x}}|
\item Reference to symbol in child scope like:  \lstinline|{x = {y = 1} z = y@x}|
\end{itemize}

\subsection{Evaluation}

To evaluate a program written in the extended language for data we follow these steps:
\begin{enumerate}
\item Parse input written in the extended language for data. This results in a tree graph.
\item Collapse references to the expressions they refer to. This results in another tree graph.
\item Serialize the evaluated tree to get the result written in the minimal language for data.
\end{enumerate}
We parse the input into a tree. In this tree a parent keeps references to its children, and each child also keeps a reference to its environment, which can be different from its parent. We use this when looking up references and collapsing them.

\section{Minimal Language for Algorithms}

We will now extend our language to handle functions. We extend the abstract syntax tree with these nodes:
\begin{itemize}
\item Conditionals like: \lstinline|if x then y else z|
\item Functions like: \lstinline|in x out (x x)|
\item Reference to function input like: \lstinline|in x out (x x)|
\item Function application like: \lstinline|{f = in x out (x x) y = f!1}|
\end{itemize}

\subsection{Evaluation}

We evaluate the program by transforming the AST to another graph. We keep a copy of the original AST. Functions know about their environment in the graph.
\begin{enumerate}
\item We do not collapse: numbers, strings, booleans
\item We collapse conditionals.
\item We collapse references.
\item We collapse function calls. When collapsing a function call we evaluate the body of the function it refers to, given the input to the function.
\item We do NOT collapse functions and do NOT collapse anything in the function body.
\end{enumerate}

\section{Implementation}

\subsection{Abstract Syntax Tree}
The parsing takes a string and builds an Abstract Syntax Tree.

\subsection{Evaluation Tree}

The evaluation takes an Abstact Syntax Tree and builds an Evaluation Tree. In the evaluation tree each node can have a pointer to its environment Node, which can be different from its parent node, e.g. for functions. The nodes in the Abstract Syntax Tree are cloned during the evaluation and given environment pointers.

In the evaluation tree each node is owned by its parent. The environment pointers are raw pointers. We do not allow a node to survive its environment. Therefore we do not allow functions to survive their bound environment, i.e. to be the output.

\subsection{Methods}

Each Node provides an evaluation method:
\begin{verbatim}
shared_ptr<Node> Expression::evaluate(const Node* environment);
\end{verbatim}
Each Node provides a lookup method that search for a reference in the node and if nothing is find passes the query to its environment.
\begin{verbatim}
shared_ptr<Node> Expression::lookup(string reference);
\end{verbatim}
Functions provide an a method for function application:

TODO: Check implementation of apply:
\begin{verbatim}
shared_ptr<Node> Expression::apply(shared_ptr<Node> input);
\end{verbatim}
How does the evaluation of the function body know about its environment?
The environment for a function body is not only its parent, but also its input.

\clearpage

\subsection{Lookup rules}

\paragraph{General Fallback}
Delegate lookup to environment. If the expression does not have an environment then report run-time error for missing symbol.

\paragraph{Dictionary}
Try lookup for each dictionary element. If the name is not found in any of the dictionary elemenets then fallback to general lookup.

\paragraph{Named Element}
If the lookup name corresponds to the name of the element then return its expression. Otherwise return null.

\paragraph{List}
If lookup name is ``first'' or ``rest'' then return that part of the list. Otherwise?

TODO: think about child lookup vs environment lookup. Will child lookup fallback to general lookup in environment? Can we have the same lookup interface for both cases or should we have two lookup interfaces?

\subsection{Evaluation Rules}
During evaluation we pass the environment to each expression that should be evaluated. During evaluation we also set the environment when creating new expressions.

\paragraph{Number} is evaluated to itself. Has no environment.
\paragraph{Character} is evaluated to itself. Has no environment.
\paragraph{String} is evaluated to itself. Has no environment.
\paragraph{Function} is evaluated to a copy of itself. Environment is set to its parent.
\paragraph{List}
\begin{enumerate}
\item is evaluated to a new list where we have evaluated each of its elements. Environment is set to its parent.
\end{enumerate}

\paragraph{Dictionary}
\begin{enumerate}
\item Create a new dictionary.
\item Evaluate each element, with an environment set to its parent extended with the partially evaluated dictionary.
\item Set environment of new dictionary to its parent.
\end{enumerate}

\paragraph{Conditional}
\begin{enumerate}
\item Evaluate the condition.
\item Delegate evaluation to either the if-child or else-child.
\end{enumerate}

\paragraph{Child reference}
\begin{enumerate}
\item Evaluate the child.
\item Look up the name in evaluated child.
\end{enumerate}

\paragraph{Reference}
\begin{enumerate}
\item Look up the name in the environment.
\item The found expreession is returned. Note that the environment of the node found is not changed.
\end{enumerate}

\paragraph{Function application}
\begin{enumerate}
\item Evaluate the input.
\item Look up the name of the function in the environment.
\item Evaluate the body of the function in its environment, extended with the input.
\end{enumerate}

\end{document}
